---
title: "CTA Ridership — Initial Exploration"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(skimr)
```

## Load

```{r load}
cta <- read_csv(
  "../data/raw/cta-data.csv",
  col_types = cols(
    station_id  = col_integer(),
    stationname = col_character(),
    date        = col_character(),
    daytype     = col_factor(levels = c("W", "A", "U")),
    rides       = col_number()
  )
) |>
  mutate(date = mdy(date))

glimpse(cta)
```

## Shape & coverage

```{r shape}
cta |>
  summarise(
    n_rows       = n(),
    n_stations   = n_distinct(station_id),
    date_min     = min(date, na.rm = TRUE),
    date_max     = max(date, na.rm = TRUE),
    n_missing    = sum(is.na(rides))
  )
```

## Overall summary

```{r skim}
skim(cta)
```

## Rides by day type

`W` = weekday, `A` = Saturday, `U` = Sunday/holiday.

```{r daytype}
cta |>
  group_by(daytype) |>
  summarise(
    n        = n(),
    mean     = mean(rides),
    median   = median(rides),
    sd       = sd(rides),
    p10      = quantile(rides, 0.10),
    p90      = quantile(rides, 0.90)
  )
```

## Busiest stations (by mean daily rides)

```{r top-stations}
cta |>
  group_by(station_id, stationname) |>
  summarise(mean_rides = mean(rides), n_days = n(), .groups = "drop") |>
  arrange(desc(mean_rides)) |>
  slice_head(n = 15)
```

## System-wide rides over time

```{r system-ts}
#| fig-width: 9
#| fig-height: 4
cta |>
  group_by(date) |>
  summarise(total = sum(rides), .groups = "drop") |>
  ggplot(aes(date, total)) +
  geom_line(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  labs(x = NULL, y = "System-wide daily rides", title = "CTA total daily ridership")
```

## Distribution of rides

```{r hist}
#| fig-width: 7
#| fig-height: 4
cta |>
  ggplot(aes(rides)) +
  geom_histogram(bins = 60) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "Rides per station-day (log10)", y = "Count")
```

## Collapse duplicate station-days

The raw file has a small number of duplicated `(station_id, date)` rows (clustered in summer 2011). We collapse them by mean so the pivot has a unique key.

```{r dedup}
dup_count <- cta |>
  count(station_id, date) |>
  filter(n > 1) |>
  nrow()

cat("duplicate (station_id, date) pairs collapsed:", dup_count, "\n")

cta_dedup <- cta |>
  group_by(station_id, stationname, date, daytype) |>
  summarise(rides = mean(rides), .groups = "drop")
```

## Reshape to wide (stations as columns)

One row per date, one column per station. Using `station_id` for column names keeps them syntactically clean; a lookup table preserves the human-readable names.

```{r wide}
station_lookup <- cta_dedup |>
  distinct(station_id, stationname) |>
  arrange(station_id)

cta_wide <- cta_dedup |>
  select(date, station_id, rides) |>
  pivot_wider(
    names_from   = station_id,
    values_from  = rides,
    names_prefix = "s_",
    names_sort   = TRUE
  ) |>
  arrange(date)

dim(cta_wide)
cta_wide[1:5, 1:6]
```

### Missingness per station (wide form)

```{r wide-missing}
cta_wide |>
  summarise(across(starts_with("s_"), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "station_id", values_to = "n_missing") |>
  mutate(station_id = as.integer(sub("^s_", "", station_id))) |>
  left_join(station_lookup, by = "station_id") |>
  arrange(desc(n_missing)) |>
  slice_head(n = 15)
```

## Write cleaned data

```{r write}
dir.create("../data/clean", showWarnings = FALSE, recursive = TRUE)
write_csv(cta,            "../data/clean/cta-data.csv")
write_csv(cta_wide,       "../data/clean/cta-data-wide.csv")
write_csv(station_lookup, "../data/clean/station-lookup.csv")
```

