---
title: "CTA Ridership — Data Cleaning"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(tseries)
library(igraph)
```

## Load ridership data

```{r load-ridership}
cta <- read_csv(
  "../data/raw/cta/cta-data.csv",
  col_types = cols(
    station_id  = col_integer(),
    stationname = col_character(),
    date        = col_character(),
    daytype     = col_factor(levels = c("W", "A", "U")),
    rides       = col_number()
  )
) |>
  mutate(date = mdy(date))
```

## Load GTFS and build adjacency matrix

Source: <https://www.transitchicago.com/developers/gtfs/>

```{r load-gtfs}
routes     <- read_csv("../data/raw/cta/routes.txt")
trips      <- read_csv("../data/raw/cta/trips.txt")
stop_times <- read_csv("../data/raw/cta/stop_times.txt")
stops      <- read_csv("../data/raw/cta/stops.txt")
```

```{r build-adjacency}
rail_routes <- routes |>
  filter(route_type == 1) |>
  select(route_id, route_long_name)

rail_trips <- trips |>
  semi_join(rail_routes, by = "route_id") |>
  group_by(route_id, direction_id) |>
  slice_min(trip_id, n = 1) |>
  ungroup() |>
  select(route_id, trip_id, direction_id, direction)

rail_stop_times <- stop_times |>
  semi_join(rail_trips, by = "trip_id") |>
  arrange(trip_id, stop_sequence) |>
  select(trip_id, stop_sequence, stop_id)

stop_parents <- stops |>
  filter(location_type == 0, !is.na(parent_station)) |>
  select(stop_id, parent_station)

rail_stations <- rail_stop_times |>
  left_join(stop_parents, by = "stop_id") |>
  left_join(rail_trips |> select(trip_id, route_id), by = "trip_id")

edges_all <- rail_stations |>
  arrange(trip_id, stop_sequence) |>
  group_by(trip_id) |>
  mutate(next_station = lead(parent_station)) |>
  ungroup() |>
  filter(!is.na(next_station), parent_station != next_station) |>
  select(route_id, node1 = parent_station, node2 = next_station)

edges <- edges_all |>
  mutate(
    a = pmin(node1, node2),
    b = pmax(node1, node2)
  ) |>
  distinct(a, b) |>
  rename(node1 = a, node2 = b) |>
  arrange(node1, node2)

all_nodes <- sort(unique(c(edges$node1, edges$node2)))
K <- length(all_nodes)
W <- matrix(0L, nrow = K, ncol = K,
            dimnames = list(all_nodes, all_nodes))

for (i in seq_len(nrow(edges))) {
  a <- as.character(edges$node1[i])
  b <- as.character(edges$node2[i])
  W[a, b] <- 1L
  W[b, a] <- 1L
}

cat("Adjacency matrix:", K, "x", K, "\n")
cat("Edges:", nrow(edges), "\n")
```

## Collapse duplicate station-days

```{r dedup}
cta_dedup <- cta |>
  group_by(station_id, stationname, date, daytype) |>
  summarise(rides = mean(rides), .groups = "drop")
```

## Reshape to weekly wide format

```{r wide}
station_lookup <- cta_dedup |>
  distinct(station_id, stationname) |>
  arrange(station_id)

cta_wide <- cta_dedup |>
  select(date, station_id, rides) |>
  mutate(week = floor_date(date, "week")) |>
  group_by(week, station_id) |>
  summarise(rides = mean(rides, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(
    names_from   = station_id,
    values_from  = rides,
    names_prefix = "s_",
    names_sort   = TRUE
  ) |>
  filter(week >= as.Date("2021-01-01"), week < as.Date("2025-01-01")) |>
  arrange(week) |>
  mutate(
    mon     = month(week),
    mon_feb = as.integer(mon == 2),
    mon_mar = as.integer(mon == 3),
    mon_apr = as.integer(mon == 4),
    mon_may = as.integer(mon == 5),
    mon_jun = as.integer(mon == 6),
    mon_jul = as.integer(mon == 7),
    mon_aug = as.integer(mon == 8),
    mon_sep = as.integer(mon == 9),
    mon_oct = as.integer(mon == 10),
    mon_nov = as.integer(mon == 11),
    mon_dec = as.integer(mon == 12)
  ) |>
  select(-mon)
```

## Stations with high missingness

```{r high-missing}
cta_wide |>
  summarise(across(starts_with("s_"), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "station", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  mutate(station_id = as.integer(sub("^s_", "", station))) |>
  left_join(station_lookup, by = "station_id") |>
  select(station, stationname, n_missing) |>
  arrange(desc(n_missing))
```

## Drop closed stations and bridge neighbors

Berwyn, Lawrence, and Damen-Lake are dropped due to extended closures (RPM renovation). Their neighbors are connected to maintain graph connectivity.

```{r drop-bridge}
drop_names <- c("Berwyn", "Lawrence", "Damen-Lake")
drop_ids <- station_lookup |>
  filter(str_detect(stationname, paste(drop_names, collapse = "|"))) |>
  pull(station_id) |>
  as.character()

cat("Dropping stations:\n")
station_lookup |>
  filter(station_id %in% as.integer(drop_ids))

# Bridge neighbors before removing
for (sid in drop_ids) {
  neighbors <- which(W[sid, ] == 1L)
  if (length(neighbors) >= 2) {
    neighbor_names <- names(neighbors)
    for (i in seq_along(neighbor_names)) {
      for (j in seq_along(neighbor_names)) {
        if (i < j) {
          W[neighbor_names[i], neighbor_names[j]] <- 1L
          W[neighbor_names[j], neighbor_names[i]] <- 1L
        }
      }
    }
  }
}

# Remove dropped stations from adjacency matrix
keep_nodes <- setdiff(rownames(W), drop_ids)
W <- W[keep_nodes, keep_nodes]

# Remove from ridership data
cta_wide <- cta_wide |>
  select(-any_of(paste0("s_", drop_ids)))

cat("\nAdjacency matrix after dropping:", nrow(W), "x", ncol(W), "\n")
```

## Filter to intersection of ridership and adjacency stations

```{r filter-intersect}
adj_stations <- paste0("s_", rownames(W))
all_stations <- names(cta_wide) |> str_subset("^s_")

in_rides_not_adj <- setdiff(all_stations, adj_stations)
in_adj_not_rides <- setdiff(adj_stations, all_stations)

cat("Stations in both:", length(intersect(all_stations, adj_stations)), "\n")
cat("In ridership but not adjacency:", length(in_rides_not_adj), "\n")
cat("In adjacency but not ridership:", length(in_adj_not_rides), "\n")

if (length(in_rides_not_adj) > 0) {
  cat("\n--- In ridership but not adjacency ---\n")
  station_lookup |>
    filter(station_id %in% as.integer(sub("^s_", "", in_rides_not_adj)))
}

if (length(in_adj_not_rides) > 0) {
  cat("\n--- In adjacency but not ridership ---\n")
  station_lookup |>
    filter(station_id %in% as.integer(sub("^s_", "", in_adj_not_rides)))
}

# Filter both to intersection
kept_ids <- intersect(
  as.integer(sub("^s_", "", all_stations)),
  as.integer(rownames(W))
) |> as.character()

W <- W[kept_ids, kept_ids]
cta_wide <- cta_wide |>
  select(week, starts_with("mon_"), all_of(paste0("s_", kept_ids)))

# Connectivity check
g <- graph_from_adjacency_matrix(W, mode = "undirected", diag = FALSE)
cat("\nConnected:", is_connected(g), "\n")
cat("Components:", components(g)$no, "\n")
cat("Final station count:", nrow(W), "\n")
```

## Missingness check

```{r missing}
missing_by_station <- cta_wide |>
  summarise(across(starts_with("s_"), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "station", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

zeros_by_station <- cta_wide |>
  summarise(across(starts_with("s_"), ~ sum(. == 0, na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "station", values_to = "n_zeros") |>
  filter(n_zeros > 0) |>
  arrange(desc(n_zeros))

cat("Stations with any NAs:", nrow(missing_by_station), "/",
    sum(str_starts(names(cta_wide), "s_")), "\n")
cat("Total missing cells:", sum(missing_by_station$n_missing), "\n")
cat("Stations with any zeros:", nrow(zeros_by_station), "\n")
cat("Total zero cells:", sum(zeros_by_station$n_zeros), "\n\n")

issues <- full_join(
  missing_by_station,
  zeros_by_station,
  by = "station"
) |>
  mutate(
    station_id = as.integer(sub("^s_", "", station)),
    n_missing = replace_na(n_missing, 0L),
    n_zeros   = replace_na(n_zeros, 0L)
  ) |>
  left_join(station_lookup, by = "station_id") |>
  select(station, stationname, n_missing, n_zeros) |>
  arrange(desc(n_missing + n_zeros))

if (nrow(issues) > 0) issues
```

## Impute zeros

```{r impute-zeros}
cta_wide <- cta_wide |>
  mutate(across(starts_with("s_"), ~ if_else(. == 0, 0.5, .)))
```

## Log first difference

```{r log-diff}
cta_wide <- cta_wide |>
  mutate(across(starts_with("s_"), ~ log(.x) - lag(log(.x)))) |>
  slice(-1)
```

## Stationarity assessment

### Time series plots

```{r ts-plots}
#| fig-width: 9
#| fig-height: 6
sample_stations <- names(cta_wide) |>
  str_subset("^s_") |>
  sample(6)

cta_wide |>
  select(week, all_of(sample_stations)) |>
  pivot_longer(-week, names_to = "station", values_to = "log_diff") |>
  ggplot(aes(week, log_diff)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~station, scales = "free_y") +
  labs(x = NULL, y = "Log first difference", title = "Sample station series")
```

## Write cleaned data

```{r write}
cta_adjacency <- as.data.frame(W) |>
  rownames_to_column("station_id")

dir.create("../data/clean", showWarnings = FALSE, recursive = TRUE)
write_csv(cta_wide,       "../data/clean/cta-data-wide.csv")
write_csv(cta_adjacency,  "../data/clean/cta-adjacency.csv")
write_csv(station_lookup, "../data/clean/station-lookup.csv")
```
