---
title: "CTA L Station Adjacency from GTFS"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

Build an edge list for the CTA L station network from the official GTFS feed.
Two stations are adjacent if they are consecutive stops on any L line.
The output is a CSV with columns `node1, node2` (station IDs) suitable for
an ICAR prior in Stan.

Source: <https://www.transitchicago.com/developers/gtfs/>

```{r setup}
library(tidyverse)
```

## Load GTFS tables

```{r load}
routes     <- read_csv("../data/raw/cta/routes.txt")
trips      <- read_csv("../data/raw/cta/trips.txt")
stop_times <- read_csv("../data/raw/cta/stop_times.txt")
stops      <- read_csv("../data/raw/cta/stops.txt")
```

## Identify rail routes

GTFS `route_type == 1` denotes rail (metro/subway).

```{r rail-routes}
rail_routes <- routes |>
  filter(route_type == 1) |>
  select(route_id, route_long_name)

rail_routes
```

## Pick one representative trip per route + direction

We only need the stop sequence, so one trip per (route, direction) is enough.
We take the first `trip_id` alphabetically for reproducibility.

```{r rep-trips}
rail_trips <- trips |>
  semi_join(rail_routes, by = "route_id") |>
  group_by(route_id, direction_id) |>
  slice_min(trip_id, n = 1) |>
  ungroup() |>
  select(route_id, trip_id, direction_id, direction)

rail_trips
```

## Get ordered stop sequences

```{r stop-seqs}
rail_stop_times <- stop_times |>
  semi_join(rail_trips, by = "trip_id") |>
  arrange(trip_id, stop_sequence) |>
  select(trip_id, stop_sequence, stop_id)

rail_stop_times
```

## Map platform stops to parent stations

Each L platform stop (30xxx) has a `parent_station` that matches the
station IDs (40xxx/41xxx) used in the ridership data.

```{r map-parents}
stop_parents <- stops |>
  filter(location_type == 0, !is.na(parent_station)) |>
  select(stop_id, parent_station)

rail_stations <- rail_stop_times |>
  left_join(stop_parents, by = "stop_id") |>
  left_join(rail_trips |> select(trip_id, route_id), by = "trip_id")

# Check for unmapped stops
unmapped <- rail_stations |> filter(is.na(parent_station))
if (nrow(unmapped) > 0) {
  warning("Unmapped stops found:")
  print(unmapped)
}

rail_stations
```

## Build edge list from consecutive stops

For each trip, consecutive parent stations form an edge. We deduplicate
across all lines so each physical edge appears only once, regardless of
how many lines share it.

```{r edges}
edges_all <- rail_stations |>
  arrange(trip_id, stop_sequence) |>
  group_by(trip_id) |>
  mutate(next_station = lead(parent_station)) |>
  ungroup() |>
  filter(!is.na(next_station), parent_station != next_station) |>
  select(route_id, node1 = parent_station, node2 = next_station)

# Normalise edge direction (smaller ID first) and deduplicate
edges <- edges_all |>
  mutate(
    a = pmin(node1, node2),
    b = pmax(node1, node2)
  ) |>
  distinct(a, b) |>
  rename(node1 = a, node2 = b) |>
  arrange(node1, node2)

cat("Total unique edges:", nrow(edges), "\n")
edges
```

## Summarise the graph

```{r summary}
station_lookup <- read_csv("../data/clean/station-lookup.csv")

all_nodes <- sort(unique(c(edges$node1, edges$node2)))
cat("Stations in adjacency graph:", length(all_nodes), "\n")

# Degree distribution
degree <- tibble(station_id = all_nodes) |>
  mutate(
    degree = map_int(station_id, ~ {
      sum(edges$node1 == .x) + sum(edges$node2 == .x)
    })
  ) |>
  left_join(station_lookup, by = "station_id") |>
  arrange(desc(degree))

degree |> print(n = 20)

cat("\nDegree distribution:\n")
table(degree$degree)
```

## Visualise which lines contribute edges

```{r line-edges}
line_edge_counts <- edges_all |>
  mutate(
    a = pmin(node1, node2),
    b = pmax(node1, node2)
  ) |>
  distinct(route_id, a, b) |>
  count(route_id, name = "n_edges") |>
  left_join(rail_routes, by = "route_id")

line_edge_counts
```

## Full adjacency matrix

```{r adjacency-matrix}
K <- length(all_nodes)
W <- matrix(0L, nrow = K, ncol = K,
            dimnames = list(all_nodes, all_nodes))

for (i in seq_len(nrow(edges))) {
  a <- as.character(edges$node1[i])
  b <- as.character(edges$node2[i])
  W[a, b] <- 1L
  W[b, a] <- 1L
}

cat("Adjacency matrix:", K, "x", K, "\n")
cat("Non-zero entries:", sum(W), "(should be", 2 * nrow(edges), ")\n")
W
```

## Write edge list

```{r write}
cta_adjacency <- as.data.frame(W) |>
  rownames_to_column("station_id")

dir.create("../data/clean", showWarnings = FALSE, recursive = TRUE)
write_csv(cta_adjacency, "../data/clean/cta-adjacency.csv")
```
