---
title: "CTA Ridership — Initial Exploration"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(skimr)
```

## Load

```{r load}
cta <- read_csv(
  "../data/raw/cta/cta-data.csv",
  col_types = cols(
    station_id  = col_integer(),
    stationname = col_character(),
    date        = col_character(),
    daytype     = col_factor(levels = c("W", "A", "U")),
    rides       = col_number()
  )
) |>
  mutate(date = mdy(date))

glimpse(cta)
```

## Shape & coverage

```{r shape}
cta |>
  summarise(
    n_rows       = n(),
    n_stations   = n_distinct(station_id),
    date_min     = min(date, na.rm = TRUE),
    date_max     = max(date, na.rm = TRUE),
    n_missing    = sum(is.na(rides))
  )
```

## Overall summary

```{r skim}
skim(cta)
```

## Rides by day type

`W` = weekday, `A` = Saturday, `U` = Sunday/holiday.

```{r daytype}
cta |>
  group_by(daytype) |>
  summarise(
    n        = n(),
    mean     = mean(rides),
    median   = median(rides),
    sd       = sd(rides),
    p10      = quantile(rides, 0.10),
    p90      = quantile(rides, 0.90)
  )
```

## Busiest stations (by mean daily rides)

```{r top-stations}
cta |>
  group_by(station_id, stationname) |>
  summarise(mean_rides = mean(rides), n_days = n(), .groups = "drop") |>
  arrange(desc(mean_rides)) |>
  slice_head(n = 15)
```

## System-wide rides over time

```{r system-ts}
#| fig-width: 9
#| fig-height: 4
cta |>
  group_by(date) |>
  summarise(total = sum(rides), .groups = "drop") |>
  ggplot(aes(date, total)) +
  geom_line(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  labs(x = NULL, y = "System-wide daily rides", title = "CTA total daily ridership")
```

## Distribution of rides

```{r hist}
#| fig-width: 7
#| fig-height: 4
cta |>
  ggplot(aes(rides)) +
  geom_histogram(bins = 60) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "Rides per station-day (log10)", y = "Count")
```

## Collapse duplicate station-days

The raw file has a small number of duplicated `(station_id, date)` rows (clustered in summer 2011). We collapse them by mean so the pivot has a unique key.

```{r dedup}
dup_count <- cta |>
  count(station_id, date) |>
  filter(n > 1) |>
  nrow()

cat("duplicate (station_id, date) pairs collapsed:", dup_count, "\n")

cta_dedup <- cta |>
  group_by(station_id, stationname, date, daytype) |>
  summarise(rides = mean(rides), .groups = "drop")
```

## Reshape to wide (stations as columns)

One row per date, one column per station. Using `station_id` for column names keeps them syntactically clean; a lookup table preserves the human-readable names.

```{r wide}
station_lookup <- cta_dedup |>
  distinct(station_id, stationname) |>
  arrange(station_id)

cta_wide <- cta_dedup |>
  select(date, station_id, rides) |>
  pivot_wider(
    names_from   = station_id,
    values_from  = rides,
    names_prefix = "s_",
    names_sort   = TRUE
  ) |>
  arrange(date)

dim(cta_wide)
cta_wide[1:5, 1:6]
```

### Missingness per station (wide form)

```{r wide-missing}
cta_wide |>
  summarise(across(starts_with("s_"), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "station_id", values_to = "n_missing") |>
  mutate(station_id = as.integer(sub("^s_", "", station_id))) |>
  left_join(station_lookup, by = "station_id") |>
  arrange(desc(n_missing)) |>
  slice_head(n = 15)
```

## Station correlation matrix

How do stations' daily ridership move together? We compute pairwise correlations across the wide-form station columns, using pairwise-complete observations to handle missing days.

```{r cor-compute}
station_mat <- cta_wide |>
  select(starts_with("s_")) |>
  as.matrix()

cor_mat <- cor(station_mat, use = "pairwise.complete.obs")

# Drop stations that produce any NA correlations (insufficient overlap with others)
keep <- !apply(cor_mat, 1, anyNA)
cat("stations dropped due to NA correlations:", sum(!keep), "\n")
cor_mat <- cor_mat[keep, keep]

dim(cor_mat)
summary(cor_mat[lower.tri(cor_mat)])
```

### Heatmap (hierarchically clustered)

Ordering rows/columns by hierarchical clustering groups stations whose ridership patterns move together.

```{r cor-heatmap}
#| fig-width: 9
#| fig-height: 8
hc_order <- hclust(as.dist(1 - cor_mat))$order
cor_ordered <- cor_mat[hc_order, hc_order]

cor_long <- as_tibble(cor_ordered, rownames = "station_a") |>
  pivot_longer(-station_a, names_to = "station_b", values_to = "cor") |>
  mutate(
    station_a = factor(station_a, levels = rownames(cor_ordered)),
    station_b = factor(station_b, levels = colnames(cor_ordered))
  )

ggplot(cor_long, aes(station_a, station_b, fill = cor)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "Pearson r",
       title = "CTA station ridership correlation (clustered)") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```

### Most- and least-correlated station pairs

```{r cor-pairs}
cor_pairs <- cor_mat
cor_pairs[upper.tri(cor_pairs, diag = TRUE)] <- NA

pair_tbl <- as_tibble(cor_pairs, rownames = "a") |>
  pivot_longer(-a, names_to = "b", values_to = "cor") |>
  drop_na(cor) |>
  mutate(
    a_id = as.integer(sub("^s_", "", a)),
    b_id = as.integer(sub("^s_", "", b))
  ) |>
  left_join(station_lookup, by = c("a_id" = "station_id")) |>
  rename(a_name = stationname) |>
  left_join(station_lookup, by = c("b_id" = "station_id")) |>
  rename(b_name = stationname) |>
  select(a_name, b_name, cor)

# Top 10 most correlated
pair_tbl |> arrange(desc(cor)) |> slice_head(n = 10)

# Bottom 10 (least / most negatively correlated)
pair_tbl |> arrange(cor) |> slice_head(n = 10)
```

## Load data from cta-station-map data and extract red line stations

```{r}
cta_station_map <- read_csv("../data/raw/cta/cta-station-map.csv") %>%
  group_by(MAP_ID) %>%
  summarise(across(everything(), any)) %>%
  select(MAP_ID, ADA:O)

keep_cols <- cta_station_map %>%
  filter(RED == TRUE)
cta_wide_red <- cta_wide %>% 
  select(date, all_of(paste0("s_", keep_cols$MAP_ID)))

cta_wide_red_limited <- cta_wide_red %>%
  select(date, s_41170, s_40910, s_40990, s_41660)

cta_wide_red_limited_mat <- cta_wide_red_limited |>
  select(starts_with("s_")) |>
  as.matrix()

cor_mat <- cor(cta_wide_red_limited_mat, method = "pearson", use = "pairwise.complete.obs")

cor_mat
```

## Write cleaned data

```{r write}
dir.create("../data/clean", showWarnings = FALSE, recursive = TRUE)
write_csv(cta,            "../data/clean/cta-data.csv")
write_csv(cta_wide,       "../data/clean/cta-data-wide.csv")
write_csv(station_lookup, "../data/clean/station-lookup.csv")
```

