---
title: "Indego Ridership — Initial Cleaning"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
```

## Load trip data

Glob all quarterly trip CSVs in `../data/raw/` (excluding station files). Each file is read with `read_indego()`, which normalizes column names and harmonizes a few schema differences across years (`start_station_id` → `start_station`, missing `bike_type` added as NA). All columns are read as character up front so heterogeneous types across files don't break `bind_rows()`; typing happens in the cleaning step.

```{r load-data}
# Setting data directory
data_dir <- "../data/raw/indego"

# Loading data file string
csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE) |>
  str_subset("station", negate = TRUE)

# Create function to load in data files
read_indego <- function(path) {
  df <- read_csv(path, show_col_types = FALSE, col_types = cols(.default = "c"))
  names(df) <- names(df) |>
    str_to_lower() |>
    str_replace_all("\\s+", "_")
  if ("start_station_id" %in% names(df)) {
    df <- df |> rename(start_station = start_station_id, end_station = end_station_id)
  }
  if (!"bike_type" %in% names(df)) {
    df <- df |> mutate(bike_type = NA_character_)
  }
  df
}

# Load in data files
trips_raw <- map(csv_files, possibly(read_indego, otherwise = NULL)) |>
  compact() |>
  bind_rows()
```

## Clean trip data

Parses start/end times across multiple date formats seen in the source files, converts pre-2017 Q2 durations from seconds to minutes, coerces numeric columns, derives year/quarter fields, applies sanity filters, and dedupes by `trip_id`.

```{r clean-data}
# Cleaning data
trips <- trips_raw |>
  mutate(
    # Making time data consistent
    start_time = suppressWarnings(coalesce(mdy_hm(start_time), mdy_hms(start_time),
                                           ymd_hms(start_time), ymd_hm(start_time))),
    end_time = suppressWarnings(coalesce(mdy_hm(end_time), mdy_hms(end_time),
                                         ymd_hms(end_time), ymd_hm(end_time))),
    duration = as.numeric(duration),
    # Pre-2017 Q2 files recorded duration in seconds; convert to minutes
    duration = if_else(start_time < ymd("2017-04-01"), duration / 60, duration),
    # Make sure numeric columns are, in fact, numeric
    start_lat = as.numeric(start_lat),
    start_lon = as.numeric(start_lon),
    end_lat = as.numeric(end_lat),
    end_lon = as.numeric(end_lon),
    plan_duration = as.numeric(plan_duration),
    # Create new time variables
    year = year(start_time),
    quarter = quarter(start_time),
    year_quarter = paste0(year, " Q", quarter)
  ) |>
  filter(
    !is.na(start_time), # Filter NA start times
    !is.na(duration), # Filter NA duration times
    duration > 0, # Make sure duration is > 0
    duration < 1440, # Filter out trips of 24 hours or greater
    year >= 2015 # Make sure year is 2015 and on (shouldn't be a problem)
  ) |>
  distinct(trip_id, .keep_all = TRUE) # De-duplicate trip IDs
```

## Load station data

```{r load-stations}
# Load station data
stations <- read_csv("../data/raw/indego/indego-stations-2026-01-01.csv", show_col_types = FALSE)
names(stations) <- names(stations) |> str_to_lower() |> str_replace_all("\\s+", "_")
stations <- stations |> mutate(station_id = as.character(station_id))
```

## Reshape to wide (stations as columns, half-hours as rows)

One row per half-hour, one column per station, values are trip counts starting at that station in that interval. Missing cells are filled with 0, since absence of a trip record means zero trips.

```{r wide-halfhour}
# Drop trips with missing / blank / literal-"NA" start_station so no s_NA column is created
trips_halfhour <- trips |>
  filter(!is.na(start_station), nzchar(start_station), start_station != "NA") |>
  mutate(half_hour = floor_date(start_time, unit = "30 minutes")) |>
  count(half_hour, start_station, name = "trips")

# Pivot wide: half_hour rows, station columns (s_ prefix keeps names syntactic)
trips_wide <- trips_halfhour |>
  pivot_wider(
    names_from   = start_station,
    values_from  = trips,
    names_prefix = "s_",
    names_sort   = TRUE,
    values_fill  = 0
  )

# Build full sequence of half-hours from earliest to latest observed trip
full_halfhours <- tibble(
  half_hour = seq(
    floor_date(min(trips$start_time, na.rm = TRUE), "30 minutes"),
    floor_date(max(trips$start_time, na.rm = TRUE), "30 minutes"),
    by = "30 min"
  )
)

# Right-join so every half-hour has a row; fill in 0s for half-hours with no recorded trips
trips_wide <- full_halfhours |>
  left_join(trips_wide, by = "half_hour") |>
  mutate(across(starts_with("s_"), ~ replace_na(., 0))) |>
  arrange(half_hour)

trips_wide <- trips_wide[, -2] #Drop NA column (not sure why it kept appearing)

dim(trips_wide)
trips_wide[1:5, 1:6]
```

```{r cor-compute}
station_mat <- trips_wide |>
  select(starts_with("s_")) |>
  as.matrix()

cor_mat <- cor(station_mat, use = "pairwise.complete.obs")

# Drop stations that produce any NA correlations (insufficient overlap with others)
keep <- !apply(cor_mat, 1, anyNA)
cat("stations dropped due to NA correlations:", sum(!keep), "\n")
cor_mat <- cor_mat[keep, keep]

dim(cor_mat)
summary(cor_mat[lower.tri(cor_mat)])
```

```{r cor-heatmap}
#| fig-width: 9
#| fig-height: 8
hc_order <- hclust(as.dist(1 - cor_mat))$order
cor_ordered <- cor_mat[hc_order, hc_order]

cor_long <- as_tibble(cor_ordered, rownames = "station_a") |>
  pivot_longer(-station_a, names_to = "station_b", values_to = "cor") |>
  mutate(
    station_a = factor(station_a, levels = rownames(cor_ordered)),
    station_b = factor(station_b, levels = colnames(cor_ordered))
  )

ggplot(cor_long, aes(station_a, station_b, fill = cor)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "Pearson r",
       title = "Indego station ridership correlation (clustered)") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```